When environmentally persistent pathogens transform good habitat into ecological traps

### 2015-07-15 Manuscript/manuscript.tex
Added sections required by Interface

@@

@@

The parameterization above assumes that all patches are equally accessible from all other patches (i.e. are separated by effective distance 1).  To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).

@@

From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (i.e. high quality habitat), but Strasser et al. (\cite{Strasser2010}) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial, and thus should be conserved.

@@

\section{Acknowledgements}

CBL would like to thank members of the Webb Lab, past and present, for several rounds of useful feedback.  This work was supported by USGS Cooperative Agreement 101485, and NSF Graduate Research Fellowship, No. DGE-1321845.

Any mention of trade, product, or firm names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

@@

%\begin{table}[h!]   
%\caption{Parameters of the SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate the same value as the default parameterization.}
%\begin{tabular}{p{1cm} p{4cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm}}
%Param. & Interpretation &  Default & Local movement & High env. stoch. & Dominant reservoir trans. & High virulence \\
%\hline
%$\xi_{im}$ & Effect of patch quality on immigration & 0.5 & -- & -- & -- & --\\
%$\xi_{em}$ & Effect of population size on emigration & 0.5 & -- & -- & --& --\\
%$D$& Inverse of mean dispersal distance & 5 & 2 & -- & -- & --\\
%$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & -- & --\\
%\hline
%$\mu$ & Extinction rate of unit quality patch & 0.1 & -- & -- & -- & --\\
%$\nu$ & Disease-induced reduction in population size & 0.2 & -- & -- & -- & 1\\
%$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & -- & --\\
%\hline
%$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1 & --\\
%$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & -- & --
%\end{tabular}
%\label{params}
%\end{table}

>>>>

\newpage

@@

\subsubsection*{Keywords:}
disease; metapopulation; habitat quality; environmental transmission; ecological trap

@@

The parameterization above assumes that all patches are equally accessible from all other patches (i.e. are separated by effective distance 1).  To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighbouring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).

@@

From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favourable for the host (i.e. high quality habitat), but Strasser et al. (\cite{Strasser2010}) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial, and thus should be conserved.

@@

\subsection*{Data accessibility}
Simulation source code and generated data are available at \url{https://github.com/clint-leach/Metapop-Disease}, along with scripts used to generate figures.

\subsection*{Competing interests}
We have no competing interests.

\subsection*{Author's contributions}
CBL participated in the design of the study, wrote the code and carried out the simulations, and drafted the manuscript; CTW and PCC conceived of the study, participated in the design of the study, and helped draft the manuscript.  All authors gave final approval for publication.

\subsection*{Acknowledgements}

CBL would like to thank members of the Webb Lab at Colorado State University, past and present, for several rounds of useful feedback.

\subsection*{Funding}

This work was supported by USGS Cooperative Agreement 101485, and NSF Graduate Research Fellowship, No. DGE-1321845.
\\
\\
Any mention of trade, product, or firm names is for descriptive purposes only and does not imply endorsement by the U.S. Government.

@@
